Viscous Erosion in Porous Media

### 2018-03-20 docs/methods/erosion_methods.tex
touched up intro

\begin{abstract} 
We develop numerical methods to simulate the fluid-mechanical erosion of many bodies (order 50) in 2D Stokes flow. The broad aim is to simulate the erosion of a porous medium (e.g.~ground water flow) with resolution at the grain scale. Our fluid solver is based on a boundary-integral formulation of the Stokes equations with a double-layer potential. We solve the associated linear system iteratively using GMRES and the fast-multipole method (FMM) for acceleration. Since the underlying integral equation is second-kind, GMRES achieves a bounded iteration count (independent of the number of unknowns). To evolve the individual grains under shear-driven erosion, we regularize via curvature-penalization and solve the associated system using the {\thL} formulation. This framework allows numerically stable treatment of stiff terms, thus permitting large time steps. The overall accuracy of our method is exponential in space and second-order in time. The method is computationally efficient, with the main part of the solver having an $O(N)$ cost, and a one-time $O(N^2)$ computation. We benchmark single-body results against analytical predictions. In our multi-body simulations, we find that erosion leads to spontaneous channelization between bodies in near contact. The channelization is associated with a dramatic reduction in the resistance of the porous medium, much more than would be expected from the reduction in grain size alone.
\end{abstract}

@@

Erosion is a fluid-mechanical process responsible for a variety
of phenomena in geophysical and other settings.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, while stream-bed erosion is responsible for branching structures
of rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in biophysical applications, for example plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}. 
Much recent work has been devoted to understanding fundamental aspects of the fluid-structure interaction that occurs during the erosion \cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, mit-spa2016, hewett2017evolution, moore2017riemann, lachaussee2018competitive, lopez2018cfd}, and related processes like dissolution or melting \cite{Huang2015, kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion, yuan2017dynamics, claudin2017dissolution, hewett2017pear}. All of these processes involve a feedback process between how boundaries move and how flow evolves and thus can be conceptually linked  as `sculpting with flow' \cite{ristroph2018sculpting}.
% Later, we can take out a lot of these references, especially ones that are cited elsewhere in the paper. For example, Mitchell and Spagnolie and several of my papers are cited elsewhere.

@@

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation (BIE) is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  Tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly (???) ---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arclength, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a 
numerically stable fashion, e.g.~via an exponential integrator or an implicit method.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study develops numerical methods
to simulate erosion of multiply-connected two-dimensional geometries.
Since groundwater flow occurs at small spatial and velocity scales, we
assume the flow is viscous (ie.~zero Reynolds number).  By allowing for
complex multiply-connected domains, we capture the effect that
interacting bodies have on erosion, and the grains in the flow are
individually resolved rather than homogenized.  We start by validating
our method on a single circular body placed symmetrically in a CHANNEL
flow.  By artificially fixing the area of the body as a function of
time, we are able to characterize the long-time behavior of the shape
and the resulting flow and compare these results with analytical
results.  Then, we compute the vorticity of erosion in
multiply-connected geometries to better characterize erosion in porous
media.

@@

To create high-fidelity and computationally efficient numerical methods,
two fast and accurate methods are employed.  First, the fluid equations
and the erosion rate (shear stress) are efficiently solved with spectral
accuracy by using a fast-multipole-accelerated BIE method.  Given the
rate of erosion, the shape is integrated in time with second-order
accuracy.  All spatial derivatives are be computed in Fourier space,
resulting in spectral accuracy in space.  To further visualize the
effect of erosion, we compute the vorticity of the flow
(Fig.~\ref{fig:50bodies}) to visualize regions of high and low erosion.

A challenge with erosion is that it creates corners at points on the
body where the shear stress vanishes.  There is recent work for solving
BIEs in geometries with corners by using
carefully chosen local expansions around each corner~\cite{rac-ser2017,
ser-rok2016} or panel-based graded meshes~\cite{hel2011,
gil-hao-mar2014, bre2012}.  The limitation of these methods is that they
require the corner's location, a non-uniform spaced mesh, and some of
the methods have a large condition number.  One solution is to use two
meshes---one for the quadrature the other for the interface tracking.
Then, interpolation between the two grids is required, and this is the
approach taken by Ojala and Tornberg~\cite{oja-tor2015}.  Instead of
using two grids and interpolation operators, we only require the uniform
grid by slightly modify the geometry to maintain sufficient regularity.
First, we add a regularization term to the shear stress to penalize
regions of high curvature.  Second, we apply a Gaussian filter to smooth
the absolute value of the shear stress before the body shape is updated.

@@

%\todo[inline]{These are some of my thoughts on reworking this first paragraph. See if you like it.}
We begin by making two fundamental design choices: first, we restrict
ourselves to two spatial dimensions, and second the bodies are
considered immobile (i.e.~fixed to their initial locations). The
restriction to 2D allows us to simulate the erosion of many bodies
(up to $50$ in this manuscript) with high resolution, giving us the
ability to accurately characterize the erosion of a porous medium. While
BIEs are available in 3D, the extra computational cost limits the number of bodies that could be practically simulated, making porous-medium simulations less feasible. The second assumption of immobile bodies affords a simple framework to study the effects of erosion in isolation. While other effects, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Related work} Simulating eroding bodies has been studied
analytically, experimentally, and numerically.  Because of the scales
involves with geophysical flows, most of these works consider high
Reynolds number flows.  For instance, the experimental work of Moore et
al.~\cite{moo-ris-chi-zha-she2013} considers on the order of $10^4$.
For low Reynolds number flows, Mitchell and Spagnolie~\cite{mit-spa2016}
simulate the erosion of a mobile body in three dimensions, but they only
consider single body.

Mitchell and Spagnolie~\cite{mit-spa2016} formulate the traction jump as
a DLP using Lorentz Reciprocal Theorem so that additional layer
potentials have to be evaluated.  That is, they never find the density
function.

The effect of erosion on a geometry is quite rich.  For instance, it has
been hypothesized and numerically observed that the erosion leads to a
shear stress that is uniform on the body which results in a uniform
erosion rate.  The shape that has a uniform shear stress is analyzed in
three dimensions by Montenegro-Johnson and Lauga~\cite{mon-lau2015}
which is an extension of the analysis done by Pironneau~\cite{pir1973}
to find a Stokesian drag minimizing shape.  However, the effect of
erosion on a porous media is less understood.  With the methods
developed in this paper, quantities such as the total drag across a
porous region as a function of time can be computed in future work.

To simulate multiple bodies, it is essential to resolve the nonlocal
effects between the different bodies.  In this work, these interactions
are included with high-order accuracy by using an integral equation
method.  Integral equations have been used extensively to simulate
viscous fluids in complex geometries.  One particular application are
particulate flows, such as bubbles or droplets, and the classic book of
Pozrikidis~\cite{poz1992} describes BIEs for such problems.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the paper} In Section~\ref{s:formulation}, we
describe the governing equations for both the fluid and describe the
erosion model using a \thL~framework.  In Section~\ref{s:method}, the
numerical methods in both space and time are described.
Section~\ref{SingleResults} gives several numerical examples.  Finally,
concluding remarks are made in Section~\ref{s:conclusions}

@@

%%-----------------------------------------------------------------%%
% RESULTS SECTION
%%-----------------------------------------------------------------%%
\section{Results: Single-body erosion}
\label{SingleResults}
With the description of the methods complete, we now present numerical results on bodies eroding in Stokes flow. We first discuss the case of a single body, as certain features can be predicted analytically and thus used for validation. In Section \ref{MultiResults} we will discuss the erosion of multiple bodies and highlight important differences that arise compared to the single-body case.

@@

\section{Results: Multi-body erosion}
\label{MultiResults}

>>>>

\begin{abstract} 
We develop numerical methods to simulate the fluid-mechanical erosion of
many bodies (order 50) in a two-dimensional Stokes flow. The broad aim
is to simulate the erosion of a porous medium (e.g.~ground water flow)
with grain scale resolution. Our fluid solver is based on a
boundary integral formulation of the Stokes equations. We solve the
associated linear system with a fast-multipole-accelerated GMRES iteration. We use a second-kind integral equation so
that GMRES requires a bounded iteration count (independent of the number
of unknowns). To evolve the individual grains under shear-driven
erosion, we regularize via curvature-penalization and solve the
associated system using the {\thL} formulation. This framework allows
numerically stable treatment of stiff terms, thus permitting large time
steps. The overall accuracy of our method is exponential in space and
second-order in time. The method is computationally efficient, with the
fluid solver requiring $\mathcal{O}(N)$ operations per GMRES iteration, and a one-time
$\mathcal{O}(N^2)$ computation to compute the shear stress. We benchmark single-body results against analytical predictions. In our multi-body simulations, we find that erosion leads to spontaneous channelization between bodies in near contact. The channelization is associated with a dramatic reduction in the resistance of the porous medium, much more than would be expected from the reduction in grain size alone.
\end{abstract}

@@

Erosion is a fluid-mechanical process responsible for a variety
of phenomena in geophysical and other settings.  For example,
wind erosion causes sand dunes to be  longer on the windward
side~\cite{han1969}, while stream-bed erosion is responsible for branching structures
of rivers~\cite{coh-dev-sey-yi-szy-rot2015}.  Erosion is
also important in biophysical applications---plaque erosion in arteries
is key to understanding heart disease~\cite{sha2002,
gro-gij-van-fer-hat-van-yua-wen2007}, and the separation of biofilms can
be explained by erosion~\cite{pic-van-hei2000}. 
Recent work has been devoted to understanding fundamental aspects of the fluid-structure interaction that occurs during the erosion~\cite{ris-moo-chi-she-zha2012, moo-ris-chi-zha-she2013, mit-spa2016, hewett2017evolution, moore2017riemann, lachaussee2018competitive, lopez2018cfd}, and related processes like dissolution or melting~\cite{Huang2015, kondratiuk2015steadily, rycroft2016asymmetric, cohen2016erosion, yuan2017dynamics, claudin2017dissolution, hewett2017pear}. All of these processes involve a feedback process between how boundaries move and how flow evolves and thus can be conceptually linked  as `sculpting with flow'~\cite{ristroph2018sculpting}.
% Later, we can take out a lot of these references, especially ones that are cited elsewhere in the paper. For example, Mitchell and Spagnolie and several of my papers are cited elsewhere.

@@

Because of the length and velocity scales of groundwater flow, we
restrict our attention to small Reynolds number flows.  For groundwater
flow, Darcy's law is often used, but this does not resolve the pore
structure and assumes a homogeneous geometry which is often not valid.
Instead, we use the incompressible Stokes equations to represent the
flow, and the individual pores are resolved which results in a complex
multiply-connected geometry.  A boundary integral equation (BIE) is
well-suited for solving the incompressible Stokes equations in a complex
geometry since it can resolve individual grains and the non-negligible
interactions between bodies are resolved. Moreover, high-order accuracy
and computational efficiency are obtained. To link the eroding body and
fluid dynamics, the body is eroded in the normal direction at a rate
that is proportional to the magnitude of the shear
stress~\cite{ris-moo-chi-she-zha2012}.  However, tracking the interface of
fluid-structure interaction problems with a long time horizon is
particularly difficult---the mesh is often distorted and tangled by the dynamics
and regions of high curvature develop.  These two issues are resolved
by, first, introducing a tangential velocity field to maintain a grid
that is equispaced in arclength, but that does not affect the dynamics,
and, second, a curvature penalization term.  Furthermore, the frame of
reference is changed to \thL~variables so that the curvature term
appears as a diffusive term that can easily be handled in a 
numerically stable fashion, e.g.~via an exponential integrator or an implicit method.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Contributions} The present study develops numerical methods
to simulate erosion of multiply-connected two-dimensional geometries.
Since groundwater flow occurs at small spatial and velocity scales, we
assume the flow is viscous (ie.~zero Reynolds number).  By allowing for
complex multiply-connected domains, we capture the effect that
interacting bodies have on erosion, and the grains in the flow are
individually resolved rather than homogenized.  We start by validating
our method on a single circular body placed symmetrically in a pipe
flow.  By artificially fixing the area of the body as a function of
time, we are able to characterize the long-time behavior of the shape
and the resulting flow and compare these results with analytical
results.  Then, we compute the vorticity of erosion in
multiply-connected geometries to better characterize erosion in porous
media.

@@

To create high-fidelity and computationally efficient numerical methods,
two fast and accurate methods are employed.  First, the fluid equations
and the erosion rate (shear stress) are efficiently solved with spectral
accuracy by using a fast-multipole-accelerated BIE method.  Given the
rate of erosion, the shape is integrated in time with second-order
accuracy.  All spatial derivatives are computed in Fourier space,
resulting in spectral accuracy in space.  To further visualize the
effect of erosion, we compute the vorticity of the flow
(Fig.~\ref{fig:50bodies}) to visualize regions of high and low erosion.

@@

%\todo[inline]{These are some of my thoughts on reworking this first paragraph. See if you like it.}
We begin by making two fundamental design choices: first, we restrict
ourselves to two spatial dimensions, and second the bodies are
considered immobile (i.e.~fixed to their initial locations). The
restriction to two dimensions allows us to simulate the erosion of many bodies
(up to $50$ in this manuscript) with high resolution, giving us the
ability to accurately characterize the erosion of a porous medium. While
BIEs are available in three dimensions, the extra computational cost limits the number of bodies that could be practically simulated, making porous-medium simulations less feasible. The second assumption of immobile bodies affords a simple framework to study the effects of erosion in isolation. While other effects, such as sedimentation or transport, undoubtedly play a role in real applications, these effects would introduce additional complexity, both conceptually and numerically.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Related work} Simulating eroding bodies has been studied
analytically, experimentally, and numerically.  Because of the scales
involves with geophysical flows, most of these works consider high
Reynolds number flows.  For instance, the experimental work of Moore et
al.~\cite{moo-ris-chi-zha-she2013} considers on the order of $10^4$.
For low Reynolds number flows, Mitchell and Spagnolie~\cite{mit-spa2016}
use a BIE for the traction jump to simulate the erosion of a mobile body
in three dimensions, but they only consider single body.

The effect of erosion on a geometry is quite rich.  For instance, it has
been hypothesized and numerically observed that the erosion leads to a
shear stress that is uniform on the body which results in a uniform
erosion rate.  The shape that has a uniform shear stress is analyzed in
three dimensions by Montenegro-Johnson and Lauga~\cite{mon-lau2015}
which is an extension of the analysis done by Pironneau~\cite{pir1973}
to find a Stokesian drag minimizing shape.  However, the effect of
erosion on a porous media is less understood.  With the methods
developed in this paper, quantities such as the total drag across a
porous region as a function of time can be computed.

To simulate multiple bodies, it is essential to resolve the nonlocal
effects between the different bodies.  In this work, these interactions
are included with high-order accuracy by using an integral equation
method.  Integral equations have been used extensively to simulate
viscous fluids in complex geometries.  One particular application are
particulate flows, such as bubbles or droplets, and the classic book of
Pozrikidis~\cite{poz1992} describes BIEs for such problems.  For nearly
touching bodies, the quadrature associated with discretizing a BIE can
lead to instabilities of specialized quadrature is not applied.  Since
our bodies do not move, we are able to use upsampling, but future work
will require a near-singular integration scheme~\cite{qua-bir2014a,
klo-bar-gre-one2013, bar-wu-vee2015, hel-oja2008a, bea-lai2001}.

Another numerical challenge with simulating erosion is that
corners develop at points where the shear stress vanishes.  There is recent work for solving
BIEs in geometries with corners by using
carefully chosen local expansions around each corner~\cite{rac-ser2017,
ser-rok2016} or panel-based graded meshes~\cite{hel2011,
gil-hao-mar2014, bre2012}.  The limitation of these methods is that they
require the corner's location, a non-uniform spaced mesh, and some of
the methods have a large condition number.  One solution is to use two
meshes---one for the quadrature the other for the interface tracking.
Then, interpolation between the two grids is required, and this is the
approach taken by Ojala and Tornberg~\cite{oja-tor2015}.  Instead of
using two grids and interpolation operators, we only require the uniform
grid by slightly modify the geometry to maintain sufficient regularity.
First, we add a regularization term to the shear stress to penalize
regions of high curvature.  Second, we apply a Gaussian filter to smooth
the absolute value of the shear stress before the body shape is updated.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\paragraph{Outline of the paper} In Section~\ref{s:formulation}, we
describe the governing equations for both the fluid and describe the
erosion model using a \thL~framework.  In Section~\ref{s:method}, the
numerical methods in both space and time are described and
Section~\ref{s:qoi} describes how to compute the vorticity, pressure,
and drag. Section~\ref{s:SingleResults} gives several numerical examples
for a single body, and Section~\ref{s:MultiResults} presents numerical
  examples for multiply-connected domains.  Finally, concluding remarks are made in Section~\ref{s:conclusions}

@@

%%-----------------------------------------------------------------%%
% RESULTS SECTION
%%-----------------------------------------------------------------%%
\section{Results: Single-body erosion}
\label{s:SingleResults}
With the description of the methods complete, we now present numerical
results on bodies eroding in Stokes flow. We first discuss the case of a
single body, as certain features can be predicted analytically and thus
used for validation. In Section~\ref{s:MultiResults} we will discuss the erosion of multiple bodies and highlight important differences that arise compared to the single-body case.

@@

\section{Results: Multi-body erosion}
\label{s:MultiResults}
